---
title: "Supersplinatus integration"
author: "Carla Cohen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Integration of two ssp samples.

Load the required packages

```{r, messages = FALSE}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
library(patchwork)
library(tidyverse)
library(batchelor)
library(harmony)
library(lisi)
library(MAST)
library(cowplot)
options(future.globals.maxSize = 4000 * 1024^2)    # Maximum allowed total size (in bytes) of global variables identified
```

## Set up

Import the RDS files created already. 
These objects have been through the QC and filtering steps but do not contain any normalisation or clustering. 

```{r}
ssp_1 <- readRDS("/Users/davidsims/obds/R/aibst/analysis/ssp_1/RDS_objects.dir/ssp-1_after_qc.rds")
ssp_2 <- readRDS("/Users/davidsims/obds/R/aibst/analysis/ssp_2/RDS_objects.dir/ssp-2_after_qc.rds")
```

Compare the QC metrics


```{r}
p1 <- VlnPlot(ssp_1, features = c("nFeature_RNA", "nCount_RNA", "percent_mt"))
p2 <- VlnPlot(ssp_2, features = c("nFeature_RNA", "nCount_RNA", "percent_mt"))
plot_grid(p1, p2, ncol = 1)
```

## Look at merged data

First create a merged object to see how the objects compare

```{r}
ssp_merged <- merge(x = ssp_1,y = ssp_2,
                       add.cell.ids = c("ssp_1", "ssp_2"))
```

Perform SCTransform

```{r, message = FALSE, warning = FALSE}
ssp_merged <- SCTransform(object = ssp_merged,
                             vars.to.regress = "percent_mt")
```

Plot the QC metrics again

```{r}
VlnPlot(ssp_merged, features = c("nCount_RNA","nCount_SCT"), pt.size = 0, same.y.lims = T, group.by = "orig.ident")
```

Plot the variable feature plot

```{r, fig.width=8}
plot1 <- VariableFeaturePlot(ssp_merged,selection.method = "sct")
top10 <- VariableFeatures(ssp_merged)[1:10]
plot2 <- LabelPoints(plot = plot1, 
                     points = top10, 
                     repel = TRUE)
plot2
```

Perform and plot PCA

```{r}
ssp_merged <- RunPCA(ssp_merged)
ElbowPlot(ssp_merged, ndims = 50)
```

Run UMAP and visualise

```{r}
ssp_merged <- RunUMAP(ssp_merged, dims = 1:50)
DimPlot(ssp_merged, group.by = "orig.ident")
```

## Integration with Seurat

Generate a named list of the two objects to run workflow

```{r}
ssp.list <- list("ssp_1" = ssp_1, "ssp_2" = ssp_2)
ssp.list
```

Apply SCTransform across the list, select integration features and prep SCT integration
```{r}

#this applies the function to each object in the list independently. 
#this is different to above where we ran the SCTransform on the merged data, so it was performed on all the data at once.
ssp.list <- lapply(ssp.list, SCTransform, variable.features.n=3000, vars.to.regress = "percent_mt") 
features <- SelectIntegrationFeatures(ssp.list, nfeatures = 3000)
ssp.list <- PrepSCTIntegration(object.list = ssp.list, 
                                anchor.features = features)

```

Find integration anchors and perform integration

```{r}
ssp.anchors <- FindIntegrationAnchors(ssp.list,
                                       anchor.features = features,
                                       normalization.method = "SCT",
                                       reduction = "cca", #default
                                       dims = 1:50) #same as merge so that we can compare outputs

ssp.integrated <- IntegrateData(ssp.anchors,
                                   dims = 1:50,
                                   normalization.method = "SCT")
```
Run PCA on integrated data

```{r}
ssp.integrated <- RunPCA(ssp.integrated, npcs = 50)
ElbowPlot(ssp.integrated, ndims = 50)
```

Visualise PCA, how does this compare to the merged data?

```{r}
DimPlot(ssp.integrated)
DimPlot(ssp_merged, reduction = "pca")
```


Run UMAP, Find Neighbours, Find Clusters
```{r}
ssp.integrated <- RunUMAP(ssp.integrated, dims = 1:30)
ssp.integrated <- FindNeighbors(ssp.integrated, dims = 1:30)
```
> various algorithms are available, louvain is the default but you can try the others to get a more fine-scale clustering e.g. leiden

Test various clustering resolutions

```{r}
ssp.integrated_res0.2 <- FindClusters(ssp.integrated, resolution = 0.2)
ssp.integrated_res0.5 <- FindClusters(ssp.integrated, resolution = 0.5)
ssp.integrated_res0.8 <- FindClusters(ssp.integrated, resolution = 0.8)
ssp.integrated_res1.0 <- FindClusters(ssp.integrated, resolution = 1.0)
```
Plot the different resolutions

```{r}
plot_res0.2 <- UMAPPlot(ssp.integrated_res0.2)
plot_res0.5 <- UMAPPlot(ssp.integrated_res0.5)
plot_res0.8 <- UMAPPlot(ssp.integrated_res0.8)
plot_res1.0 <- UMAPPlot(ssp.integrated_res1.0)
plot_grid(plot_res0.2, plot_res0.5, plot_res0.8, plot_res1.0, labels = c("0.2", "0.5", "0.8", "1.0"))
```

```{r}
ssp.integrated <- FindClusters(ssp.integrated, resolution = 0.2)
```


Plot
```{r}
DimPlot(ssp.integrated)
DimPlot(ssp.integrated, group.by = "orig.ident")
```

Find markers

```{r}
DefaultAssay(ssp.integrated) <- "RNA"

ssp.integrated.markers.LR <- FindAllMarkers(ssp.integrated, method = "LR", 
                                  latent.vars = "orig.ident", 
                                  only.pos = TRUE)
```

Get the top 5 markers for each cluster

```{r}
ssp.integrated.markers.top5 <- ssp.integrated.markers.LR %>% 
    group_by(cluster) %>% 
    slice_max(order_by = avg_log2FC, n = 5)
head(ssp.integrated.markers.top5)
```

Plot heatmap

```{r}
DefaultAssay(ssp.integrated) <- "SCT"
DoHeatmap(ssp.integrated,
          features = ssp.integrated.markers.top5$gene)
```
Dotplot

```{r}
#get list of genes for first cluster
group0 <- ssp.integrated.markers.LR %>% 
    group_by(cluster) %>% 
    slice_max(order_by = avg_log2FC, n = 10) %>% 
    filter (cluster == "0")

DotPlot(ssp.integrated,
        features = unique(group0$gene)) +
        ggtitle("cluster 0") + 
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Key markers:

TNNT2+ TRDN+ cells: "TRDN", "DES", "TNNT3", "MYH1", 
FBLN fibroblasts: "COL1A2", "COL3A1", "DCN", "NEGR1", "FBLN1", 
TNNT1+ TRDN+ cells: "TNNT1", "ATP2A2", "MYH7", 
MKX+ fibroblasts: "PIEZO2", "COL12A1", "MKX", 
COL22A1+ TRDN cells: "COL22A1", "CPM", "SORBS2", 
Vascular endothelium: "PECAM1", "PTPRB", "FLT1", "VWF", 
Satellite cells: "PAX7", "CALCR", "GREM1", 
Mural cells: "NOTCH3", "PDGFRB", "MYO1B", 
Lymphatic endothelium: "MMRN1", "PROX1", "KDR", "FLT4", 
Smooth muscle cells: "MYH11", "NTRK3", "ITGA8", 
Adipocytes: "GPAM", "AQP7", "ADIPOQ", 
Immune cells: "PTPRC", "CD247", "CD69", "BLNK", "CD163", "MSR1"


```{r}
geneName <- c("TRDN", "DES", "TNNT3", "MYH1", "COL1A2", "COL3A1", "DCN", "NEGR1", "FBLN1", "TNNT1", "ATP2A2", "MYH7", "PIEZO2", "COL12A1", "MKX", "COL22A1", "CPM", "SORBS2", "PECAM1", "PTPRB", "FLT1", "VWF", "PAX7", "CALCR", "GREM1", "NOTCH3", "PDGFRB", "MYO1B", "MMRN1", "PROX1", "KDR", "FLT4", "MYH11", "NTRK3", "ITGA8", "GPAM", "AQP7", "ADIPOQ", "PTPRC", "CD247", "CD69", "BLNK", "CD163", "MSR1")

DotPlot(ssp_merged, features = geneName, cols = c("blue", "red")) + 
    scale_x_discrete(breaks=geneName, labels=geneName) + 
    ggtitle("Top markers") + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Save the Seurat objects

```{r}
saveRDS(ssp.integrated, "RDS_objects.dir/ssp_seurat_integrated.rds")
```

