---
title: "Example code for ggplot2"
author: "Zedias Chikwambi"
date: "10/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)

```
```{r}
install.packages("ggpubr")
library(ggpubr)
```
# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?
Answer: It lays out the plotting area.

```{r}
ggplot(diamonds)
```
#This does not allow one to see the whole graph because there are not specifics. e.g

> Answer:
The figure is just gray, with nothing plotted in the space.
## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:The axis (x&y) are ploted with labels shown

## Add geometric objects
#With geom_plot - data is plotted as points on the graph indicating the price of each carat.

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = cut)) +geom_point() + geom_smooth()
plot1<-ggplot(diamonds, aes(x = carat, y = price, color = cut)) +geom_point() + geom_smooth()
  
```
#geomesmooth - draws lines through each data set to indicate the general trend of the data (mean and sd)
- Color data points by `cut`.
# Colors the data points and splits them by cut.
  (Copy-paste and extend the code chunk above.)

```{r}
plot2<-ggplot(diamonds, aes(x = carat, y = price, color = cut)) +geom_point(aes(color = cut)) + geom_smooth()
  
```
# 
- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = cut)) +geom_point() + geom_smooth(aes(color = cut))
obds_diamonds<-ggplot(diamonds, aes(x = carat, y = price, color = cut)) +geom_point() + geom_smooth(aes(color = cut))
```

- Assign that last plot to an object called `obds_diamonds`.

```{r}
plot3<-ggplot(diamonds, aes(x = carat, y = price, color = cut)) + geom_smooth(aes(color = cut))
  
```


```{r}
plot4<-ggplot(diamonds, aes(x = carat, y = price, color = cut)) + geom_smooth()
  
```

#combine the plots(1:4) on one page

```{r}
ggarrange(plot1, plot2, plot3, plot4 + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```


```{r}
ggarrange(plot1, plot2, plot3, plot4 + rremove("x.text"), 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```
# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:plot1 provides trendlines for each cut, while plot2 provides a trendline for the global dataset irrispective of cut.

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
data(ChickWeight)#Load the data
?ChickWeight #to view the description of the data
summary(ChickWeight)
plot5<-ggplot(ChickWeight, aes(x = Time, y=weight, color = Diet)) +
  geom_point() + geom_smooth()

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
plot6<-ggplot(ChickWeight, aes(x = Time, y=weight, color = Diet)) +
  geom_point() + geom_smooth(method = lm)# draw straight lines

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)
  - Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(ChickWeight, aes(x = Time, y=weight, color = Diet)) +
  geom_point() + facet_wrap(~Diet, nrow = 2) + geom_smooth(method = lm)

```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:Diet 3 yielded bigger chickens after time (20) and 1 had the least. Diet 4 however, showed the most uniform bird growth.


- Assign that last plot to an object called `obds_chickweight`.

```{r}
obds_chickweight<-ggplot(ChickWeight, aes(x = Time, y=weight, color = Diet)) +
  geom_point() + facet_wrap(~Diet, nrow = 2) + geom_smooth(method = lm)



```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
data("msleep")
?msleep
ggplot2::msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x=order)) + geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order)) + geom_bar() + theme(axis.text.x = element_text(angle = 90))




```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x=order)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust= 1))

```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x=order)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust= 1, vjust = 0.5))


```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x=order)) + geom_bar(aes(color= genus)) + theme(axis.text.x = element_text(angle = 90, hjust= 1, vjust = 0.5, size = rel(1.5)))



```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x=order))+ 
  geom_bar(aes(fill = genus),color="black",linewidth=0.6)+ 
             theme(axis.text.x = element_text(angle = 90, hjust= 1, vjust = 0.5, size = rel(1.5)))+
             labs(title="msleep", x="order", y= "observation")+ 
                    theme(axis.title=element_text(size=16))+
                    theme(plot.title = element_text(hjust=0.5))+
                    guides(fill = guide_legend(ncol=3)) +
  theme(legend.key.size = unit(2,"mm"))
                    

```


- Assign that last plot to an object called `obds_msleep`.

```{r}
obds_msleep<-ggplot(msleep, aes(x=order))+ 
  geom_bar(aes(fill = genus),color="black",linewidth=0.6)+ 
             theme(axis.text.x = element_text(angle = 90, hjust= 1, vjust = 0.5, size = rel(1.5)))+
             labs(title="msleep", x="order", y= "observation")+ 
                    theme(axis.title=element_text(size=16))+
                    theme(plot.title = element_text(hjust=0.5))+
                    guides(fill = guide_legend(ncol=3)) +
  theme(legend.key.size = unit(2,"mm"))

```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=12, fig.width=12}
library(cowplot)
first_row <- cowplot::plot_grid(
  
  
)
second_row <- cowplot::plot_grid(
  
)
super_plot <- cowplot::plot_grid(
  
  
)
super_plot
```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)

```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)





```


```{r}
ggarrange(obds_diamonds, obds_chickweight, obds_msleep + rremove("x.text"), 
          labels = c("A", "B", "C"),
          ncol = 2, nrow = 2)
```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggplot_graphs<-ggarrange(obds_diamonds, obds_chickweight, obds_msleep + rremove("x.text"), 
          labels = c("A", "B", "C"),
          ncol = 2, nrow = 2)
ggplot_graphs
```


```{r}
ggsave(
  "graphs_ggplot.jpeg",
  plot = ggplot_graphs
)
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}




```
