---
title: "OBDS Gene Set Enrichment and Pathway Analysis Practical"
author: "Charlie George"
date: "15/10/2021"
output: html_document
---

This is the pathway enrichment analysis practical of the OBDS course

- The first part of this pracical is using gprofiler to perform ORA on output dataset of differentially expressed genes from DESeq2

- The major aim of this exercise is to get you used to reading online tool documentation and playing with different paramters.

# load packages 
```{r}
# load librarys

library(tidyverse)
library(gprofiler2)
```


# Read in our tables from this morning in a new Rmd file
```{r}
# read in results table we output from DESeq2 this morning using tidyverse read_csv()

all_genes_df <- read.csv("~/OBDS_training/pathway_analysis/data/CD8_vs_CD4_DE_results_lfcshrinkage.csv")
dim(all_genes_df)
# check its dimensions - and look at the top of the file -> what does this tell us about our table? 
# dim = 22,535 -> this contains all genes -> need to filter to get those tested 

```

We want to perform a classic Over representation analsis (ORA) using gprofiler - to do this we first need to create our gene subsets 

```{r}

# Subset 1 - All genes tested in differential expression (this is our background/control gene list) - i.e. remove padj == NA 
all_genes_tested_df <- all_genes_df[!is.na(all_genes_df$padj), ]
  
dim(all_genes_tested_df)
  
# Subset 2 - All sig upregulated genes (padj < 0.05. & log2FoldChange > 1)
upreg_genes_df <- all_genes_tested_df[all_genes_tested_df$padj< 0.05 & all_genes_tested_df$log2FoldChange> 1, ]


# Subset 3 - All sig downregulated genes (padj < 0.05. & log2FoldChange < 1)
downreg_genes_df <- all_genes_tested_df[all_genes_tested_df$padj < 0.05 & all_genes_tested_df$log2FoldChange< -1, ]


# Check the dimensions of each of your dataframes using dim()
dim(downreg_genes_df)
dim(upreg_genes_df)
# all_genes_df_filtered should be 17729
# upreg_genes_df should be 915
# downreg_genes_df should be 1204

```


```{r}
# from each of these dataframes get a vector of the ensembl ids 
all_gene_ids <- all_genes_tested_df$ensembl_gene_ids
    
upreg_gene_ids <- upreg_genes_df$ensembl_gene_ids
    
downreg_gene_ids <- downreg_genes_df$ensembl_gene_ids
```

#Perform ORA using g:profiler 

1. Look at documentation - https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
```{r}
# use the gost() function - what parameters should we include/change??
# we want to test the ORA of upreg_gene_ids vs our all_gene_ids

# get help for function
# write query and assign to output variable named `gost_results_obj`
gost_results_obj <- gost(
  query = upreg_gene_ids,
  organism = "mmusculus",
  exclude_iea = TRUE,
  correction_method = c("g_SCS"),
  custom_bg = all_gene_ids,
  )

gost_results_obj$result




# what are our outputs  


```

```{r}
# plot our results using in built function in g:profiler -> which one is it? 
gostplot(
  gostres = gost_results_obj,
  capped = TRUE,
  interactive = TRUE,
  pal = c(`GO:MF` = "#dc3912", `GO:BP` = "#ff9900", `GO:CC` = "#109618", KEGG =
    "#dd4477", REAC = "#3366cc", WP = "#0099c6", TF = "#5574a6", MIRNA = "#22aa99", HPA =
    "#6633cc", CORUM = "#66aa00", HP = "#990099")
)

```

```{r}
# what does as_short_link = FALSE do in gost() function? 
gost_results_obj_sl <- gost(
  query = upreg_gene_ids,
  organism = "mmusculus",
  exclude_iea = TRUE,
  as_short_link = TRUE,
  correction_method = c("g_SCS"),
  custom_bg = all_gene_ids,
  )
gost_results_obj_sl
```

Answer -> 

```{r}
# Filter results table for genesets with between 5-300 genes 



```


2. Try querying multiple gene lists at the same time (up and down regulated genes) using Multiple queries - see the documentation 

```{r}

```

Can you plot these querys together?

```{r}


```

Output the get_version_info() to save the versions of the databases that were used for the analysis!! 
```{r}

```

Output the sessionInfo() so know what packages we used
```{r}
sessionInfo()

```

Now knit your report to get a lovely html file that you can share -> this is good practise as it creates a permanent record of your plots/analysis output and also makes sure your script runs without error from start to finish! 